# Week 2 Asynchronous

Getting Started with Data Visualization in R > Week 2

```{r, include = FALSE}
source("_common.R")
```

## Introduction to the tidyverse

```{r eval=FALSE}
install.packages("tidyverse")
```

## Data import and structure in the tidyverse
## Filtering, selecting, recoding, renaming, and piping
## Recoding, Renaming, and Calculating Columns
## Grouping and summarizing data

```{r, code = readLines("week2/1w2_1.R")}

```

## Practices
```{r}
library(tidyverse)
```

```{r}
dat <- drop_na(read_csv(url("https://www.dropbox.com/s/uhfstf6g36ghxwp/cces_sample_coursera.csv?raw=1")))
```

```{r}
# Problem 1

# The object "dat" created in the assignment code will import the survey data for the assignment using read_csv, thereby creating a tibble. Using that object as your data, use select() to create a new tibble that include only the columns for educational level, whether the respondent has an educational loan, employment status, and Trump approval. Display that object. Hint: consult the codebook to identify the correct column names.

#### Write your code below:
dat %>% head()

df <- dat %>% select(educ, edloan, employ, CC18_308a)
df %>% head()
```

```{r}
# Problem 2

# Continuing to use the new data table you created in Problem 1, use recode() to create a new column named "trump_approve_disapprove" that recodes the column for President Trump's job approval. A value of "1" should mean that the respondent either "strongly" or "somewhat" approves of the President, and a value of 0 should mean that the respondent either "strongly" or "somewhat" DISapproves of the president. Display the resulting object. 


#### Write your code below:

```

CC18_308a Job approval by President Trump  
Do you approve or disapprove of the way each is doing their job...  
1 Strongly approve  
2 Somewhat approve  
3 Somewhat disapprove  
4 Strongly disapprove  

```{r}
df$trump_approve_disapprove <- df$CC18_308a %>% 
  recode(`1` = 1,
         `2` = 1,
         `3` = 0,
         `4` = 0)
```

```{r}
df %>% head()
```

```{r}
# Problem 3

# Use summarise() to create a summary table for survey respondents who are employed full time and are married. The table should have the mean and median for the importance of religion column.


#### Write your code below:

```

employ: Which of the following best describes your current employment status?  
	1 Full-time  
	2 Part-time  
	3 Temporarily laid off  
	4 Unemployed  
	5 Retired  
	6 Permanently disabled  
	7 Homemaker  
	8 Student  
	9 Other  

marstat: What is you marital status?  
	1 Married  
	2 Separated  
	3 Divorced  
	4 Widowed  
	5 Never married  
	6 Domestic/civil partnership  
	
pew_religimp: How important is religion in yuor life?  
	1 Very important  
	2 Somewhat important  
	3 Not too important  
	4 Not at all important  
	
```{r}
dat %>% filter(employ == 1 & marstat == 1) %>% 
  summarise(`Mean Importance of Religion` = mean(pew_religimp), 
            `Median Importance of Religion` = median(pew_religimp))
```
	
# Week 2 Synchronous
## Goal

- Warm-up
- Transform data in action
- Tidy data in action

Examples are from https://r4ds.had.co.nz/index.html

## Warm-up
### Introduce short cut

insert chunk  
option + command + I (MAC)  
control + alt + I (Windows)  

%>%  
cmd/control + shift + M  

<-  
option + - (MAC)  
alt + - (Windows)  

help in R  
?function_name

### Subsetting
```{r}
library(tidyverse)
```

```{r}
set.seed(1234)
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df
```

#### I want to extract column x.
```{r}
# Extract by name
df$x
df[["x"]]

df %>% .$x
df %>% .[["x"]]
df %>% pull(x)

# Extract by position
df[[1]]
```

## Transform data with dplyr

We will use data from the nycflights13 package.

```{r eval=FALSE}
install.packages("nycflights13")
```

After we installed it, let's import it.

```{r}
library(nycflights13)
```

Here is the data.

```{r}
flights
```

This data frame contains all 336,776 flights that departed from New York City in 2013. 

To know more about it, use help function.

```{r}
?flights
```

int stands for integers  
dbl stands for doubles, or real numbers  
chr stands for character vectors, or strings  
dttm stands for date-times (a date + a time)  

### Pick rows (observations): filter()
```{r}
?filter
```

#### What are all flights on January 1st?
```{r}
flights %>% 
  filter(month == 1, day == 1)
```

#### What are all flights on December 25th?
```{r}
flights %>% 
  filter(month == 12, day ==25)
```

#### What are all flights that departed in November or December?

Here we need some logical operators

& is “and”, | is “or”, and ! is “not”.

```{r}
flights %>% 
  filter(month == 11 | month == 12)
```

Another way is to use %in%, this can save some writing。

```{r}
flights %>% 
  filter(month %in% c(11, 12))
```

### Reorder rows: arrange()

arrange() works similarly to filter() except that instead of selecting rows, it changes their order.

#### I want the table ordered by date (year, month, day).
```{r}
flights %>% 
  arrange(year, month, day)
```

#### I want the table ordered by ascending departure delay.
```{r}
flights %>% 
  arrange(dep_delay)
```

#### I want the table ordered by descending departure delay.
```{r}
flights %>% 
  arrange(desc(dep_delay))
```

### Pick columns (variables): select()
#### I only need the column of year, month, day, and origin, dest
```{r}
flights %>% 
  select(year, month, day, origin, dest)
```

```{r}
flights %>% 
  select(year:day, origin, dest)
```

#### I want the time_hour to be the first column
```{r}
flights %>% 
  select(time_hour, everything())
```

### Make new columns (variables): mutate()

Besides selecting sets of existing columns, it’s often useful to add new columns that are functions of existing columns. That’s the job of mutate().

#### I want the speed of each flight.
```{r}
flights %>% 
  select(distance, air_time) %>% 
  mutate(speed_miles_minute = distance / air_time,
         speed_miles_hour = speed_miles_minute * 60)
```

### Collapse many values down to a single summary: summarise()
#### What is the average departure delay over the year?
```{r}
flights %>% 
  summarise(delay = mean(dep_delay))
```

```{r}
flights %>% 
  summarise(delay = mean(dep_delay, na.rm = TRUE))
```

#### What is the average departure delay for each month?

Functions will be automatically applied “by group”.

```{r}
flights %>% 
  group_by(month) %>% 
  summarise(delay_month = mean(dep_delay))
```

```{r}
flights %>% 
  group_by(month) %>% 
  summarise(delay_month = mean(dep_delay, na.rm = TRUE))
```

#### What is the number of flights for each month?
```{r}
flights %>% 
  group_by(month) %>% 
  summarise(n = n())
```

#### Exercise: What is the number of canceled flights for each month?

Missing values in dep_delay, arr_delay represent cancelled flights.

```{r}
flights %>%
  group_by(month) %>%
  summarise(all = n(),
            canceled = sum(is.na(dep_delay) & is.na(arr_delay)))
```

## Tidy data with tidyr

1. Each variable must have its own column.
1. Each observation must have its own row.
1. Each value must have its own cell.

```{r}
table1
```

### Reshape data
#### pivot_longer()

Look at table4a

- The column names 1999 and 2000 represent values of the year variable.
- The values in the 1999 and 2000 columns represent values of the cases variable, and each row represents two observations, not one.

```{r}
table4a
```

```{r}
table4a %>% 
  pivot_longer(cols = c(`1999`, `2000`),
               names_to = "year",
               values_to = "cases")
```

```{r}
table4b
```

```{r}
table4b %>% 
  pivot_longer(cols = c(`1999`, `2000`),
               names_to = "year",
               values_to = "population")
```

#### pivot_wider()

One observation is scattered across multiple rows.

Look table2: an observation is a country in a year, but each observation is spread across two rows.

```{r}
table2
```
```{r}
?pivot_wider
```

```{r}
table2 %>% 
  pivot_wider(names_from = type, 
              values_from = count)
```

They are complement

### Split cells
#### separate()
```{r}
?separate
```

table3: one column (rate) that contains two variables (cases and population).

```{r}
table3
```

separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears.

```{r}
table3 %>% 
  separate(col = rate,
           into = c("cases", "population"))
```

```{r}
table3 %>% 
  separate(col = rate,
           into = c("cases", "population"),
           sep = '/')
```

Convert chr to int

```{r}
table3 %>% 
  separate(col = rate,
           into = c("cases", "population"),
           sep = '/',
           convert = TRUE)
```

#### unite()
```{r}
?unite
```

```{r}
table5
```

```{r}
table5 %>% 
  unite(new, century, year)
```

Here we don’t want separator "_" so we use "":

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

### Missing values

A value can be missing in one of two possible ways:

Explicitly, i.e. flagged with NA.  
Implicitly, i.e. simply not present in the data.

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

stocks
```

#### drop_na()
```{r}
stocks %>% drop_na()
```

#### fill()
```{r}
?fill
```

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

treatment
```

```{r}
treatment %>% 
  fill(person)
```

### Case study

The who dataset contains tuberculosis (TB) cases broken down by year, country, age, gender, and diagnosis method. The data comes from the 2014 World Health Organization Global Tuberculosis Report.

#### Look at data
```{r}
who
```

```{r}
?who
```

```{r}
colnames(who) # look at all column names
```

#### Gather together the columns that are not variables
```{r}
who %>% 
  pivot_longer(cols = new_sp_m014:newrel_f65,
               names_to = "key",
               values_to = "cases")
```

There are a lot of missing values in the current representation, so for now we’ll use values_drop_na just so we can focus on the values that are present.

```{r}
who1 <- who %>% 
  pivot_longer(cols = new_sp_m014:newrel_f65,
               names_to = "key",
               values_to = "cases",
               values_drop_na = TRUE)

who1
```

#### Formalize `key` column
```{r}
who2 <- who1 %>% 
  mutate(key = str_replace(key, "newrel", "new_rel"))
who2
```

#### Separate `key` column
```{r}
who3 <- who2 %>% 
  separate(key, 
           c("new", "type", "sexage"), 
           sep = "_")

who3
```

#### Remove useless columns
```{r}
who4 <- who3 %>% 
  select(-c(iso2, iso3, new))
who4
```

#### Separate sex and age
```{r}
who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

```{r}
# pipeline
who %>% 
  pivot_longer(cols = new_sp_m014:newrel_f65,
               names_to = "key",
               values_to = "cases",
               values_drop_na = TRUE) %>% 
  mutate(key = str_replace(key, "newrel", "new_rel")) %>% 
  separate(key, c("new", "type", "sexage"), sep = "_") %>% 
  select(-c(iso2, iso3, new)) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```